### 2012-05-24 4.tex
back to chapter files, lots of writing

%\chapter{Accidental and structural gaps in English syllable contact\protect\footnotemark{*}\protect\footnotetext{An abridged version of this chapter will appear in the proceedings of the 30th West Coast Conference on Formal Linguistics.}}
\chapter{Structural and accidental gaps\protect\symbolfootnote[1]{An abridged version of this chapter will appear in the proceedings of the 30th West Coast Conference on Formal Linguistics under the same name.}}
\label{clusters}

@@

\input{4.1}
\input{4.2}
\input{4.3}

>>>>

\chapter{Structural and accidental gaps\protect\symbolfootnote[1]{An abridged version of this chapter will appear in the proceedings of the 30th West Coast Conference on Formal Linguistics under the same name.}}
\label{clusters}

@@

\section{Domain}

% WHY I'm DOING THIS STUDY
% describes makeup of the corpus and how p

\subsection{Data sources}

The English CELEX database \citep{CELEX},

\citet{Aronoff1976}

\citet{Harley2009}

\citet{Taft1975} ?
\citet{Taft1981} ?
\citet{Taft2004a} ?

\citet{Halle1962}

% 2.1.4: k

\subsubsection{Corpus counts}

Under the phonological analyses described in the next two sections, the 6,876 simplex words in CELEX contain 21 unique medial codas and 40 unique medial onsets. Of these 840 ($= 21 \times 40$) possible syllable contact clusters that could be produced by free combinations of the attested medials codas and onsets, 158 are attested, for a 18.8\% saturation rate.

\subsection{Syllabification}

For this task, it is necessary to separate medial consonant clusters into codas and onsets. While CELEX provides syllabified transcriptions, no description of the syllabification procedure is given in the documentation, and inspection of these syllabifications suggests the procedure used is not fully systematic. For instance, \emph{chemistry} is syllabified as [ˈkɛ.mɪ.strɪ] but \emph{ministry} as [ˈmɪ.nɪs.trɪ].\footnote{Note that word-final \emph{y} is generally lax [ɪ] in Received Pronunciation \citep[][294]{AOE2}.} This putative [ɪ.strɪ $\sim$ ɪs.trɪ] contrast, and many other such contrasts in the CELEX data, are dubious simply because there is no clear evidence that contrastive syllabification occurs in any language. Apparent counterexamples appear to be effects of vowel or consonant length \citep[e.g.,][]{Elfner2006}, word stress, or morphological structure, none of which explain the \emph{chemistry}/\emph{ministry} syllabification contrast.

An automated syllabification system was constructed and applied to medial clusters in the CELEX data. This system need only deal with simplex English words that contain the medial consonant clusters which are the focus of this study.  There is, for instance, no need to address the status of ``ambisyllabic'' consonants, i.e., singleton medial consonants preceded by a stressed lax vowel \citep[][219f.]{Rubach1996}, or to address morphological effects on syllabification. The technique used is a variation on the theme of onset maximization (e.g., \citealt{Kurylowicz1948}, \citealt[42f.]{Kahn1976}, \citealt[][?]{Selkirk1983}) which parses word-medial clusters so that the onset is as large as possible. Medial clusters in words like \emph{neu}[.tr]\emph{on} or \emph{bi}[.str]\emph{o} are found in word-initial position (e.g., [tr]\emph{ansit}, [str]\emph{ike}), and onset maximization leaves the coda of the first syllable empty. In contrast, the [nstr] cluster in \emph{mi}[n.str]\emph{el} does not occur word-initially; here the maximal onset is [str], and the [n] is assigned to the coda.

There is one case where unchecked maximization of medial onsets produces incorrect syllabifications. When a medial consonant cluster is preceded by a stressed lax vowel, as in words like \emph{propaga}[n.d]\emph{a}, \emph{whi}[s.p]\emph{er}, \emph{vi}[s.t]\emph{a}, or \emph{bi}[s.k]\emph{uit}, the first consonant of the cluster checks the lax vowel \citep[e.g.,][3]{Hammond1997}. \citet[][55]{Harris1994} notes that onset maximization makes the wrong predictions when the medial cluster is a valid onset: in \emph{whisper}, \emph{vista}, and \emph{biscuit}, it incorrectly assigns [sp, st, sk] to the medial onset, leaving the coda empty. The approach adopted here is to first apply onset maximization, then to reassign the first consonant of a medial onset to the preceding coda if it is immediately preceded by a stressed lax vowel.\footnote{\citet[][589f.]{Lowenstamm1981} observes a strikingly similar issue with the onset maximization principle in French, and proposes a similar solution.}

If the English affricates [tʃ, dʒ] are treated as sequences and not individual segments, this addition to the standard onset maximization technique would itself make the wrong prediction for medial affricates preceded by lax stressed vowels in words like \emph{ra}[.tʃ]\emph{et} or \emph{a}[.dʒ]\emph{ile}, incorrectly assigning the stop and fricative portions to separate syllables. For this reason, an additional constraint is assumed which prevents the two components of the affricates from being split by syllabification. This is motivated by the tendency of affricates to pattern with single segments in many languages. For instance, the only complex onsets in Classical Nahua are the affricates [ts, tʃ, tɬ] \citep[][9]{Launey2011}.

%In fact, [t.ʃ, d.ʒ] clusters are not found in simplex English words, despite the fact that that affricates occur as medial onsets in clusters (e.g., \emph{tru}[n.tʃ]\emph{eon}, \emph{so}[l.dʒ]\emph{er}).

\subsection{Inventory considerations}

To identify the inventory of syllable contact clusters, it is also necessary to parse syllables into onset, nucleus, and coda. In most cases this is trivial, but a few distributional heuristics for difficult cases are described below.

\subsubsection{The velar nasal}
\label{velarnasal}

There is a long-standing debate regarding whether English [ŋ] is a phoneme in its own right, as demanded by Kiparsky's Alternation Condition \citep{Kiparsky1968} or Lexicon Optimization \citep[][53]{OT}, or simply the allophone of /n/ found before /k, g/ \citep[][65]{Borowsky1986}. The strongest piece of evidence for treating the velar nasal as a pure allophone is the general absence of ?
nset [ŋ], a position where it can never be followed by a dorsal consonant needed to derive the velar allophone. \citet{Pierrehumbert1994} assumes the pure allophonic analysis in her study of syllable contact clusters, and it is assumed here. English nasal place allophony is formalized in Section \ref{cnpasection} below.
%\citealt[][62]{Halle1985a}),

\subsubsection{[j] onglides}

The [j] onglide in words such as in words such as \emph{ass}[j]\emph{ume} is assumed to be present in underlying representations (e.g., \citealt[][278]{Borowsky1986}, J. \citealt{Anderson1988b}, pace \emph{SPE}:196, \citealt[][89]{Halle1985a}, \citealt[][217]{McMahon1990}), since the presence or absence of the glide is at least marginally contrastive (e.g., \emph{coo}/\emph{coup} \alt{} \emph{queue}, \emph{booty} \alt{} \emph{beauty}). The front onglide is further assumed to be assigned to the nucleus, except when the onset would otherwise be null (e.g., \emph{jun}[j]\emph{or}). 
%\footnote{English glides are transcribed here as full segments, not as ``subsegments'', as this distinction does not appear to be meaningful for English, or empirically motivated for other languages \citep{Rubach2002}.}
There is extensive motivation for this principle. When [j] is a simplex onset, it may be followed by any vowel \citep[][276]{Borowsky1986}, but when [j] is immediately preceded by an onset consonant (e.g., [bj]\emph{ugle}), the following vowel is always [uː]. In fact, speakers judge words in which the front onglide is preceded by an onset consonant and followed by a vowel other than [uː] to be anomalous \citep{Moreland2009}. This defective distribution of vowels following [j] suggests that the onglide in this context is the first component of a phonological diphthong (\citealp[][232]{Hayes1980}, \citealp[][61f.]{Harris1994}, \citealp{Davis1995}). \citet[][42]{Clements1983} note that /m, v/ do not appear in onset clusters except in words like \emph{muse} or \emph{view}; the exceptionality of [juː] also suggests the glide is nuclear. There is also strong external support for this principle. The [juː] in words like \emph{spew} may pattern together in Pig Latin \citep{Davis1995,Idsardi2005} and \emph{shm}-reduplication \citep{Nevins2003} to the exclusion of the rest of preceding tautosyllabic consonants. The same fusion of [juː] at the expense of the onset is also found in speech errors, e.g., [kjuː]\emph{mor} [h]\emph{omponent} for [hjuː]\emph{mor component} \citep[][130]{Shattuck-Hufnagel1986}. Finally, \citet{Buchwald2005} considers [j] onglides in the speech of VBR, an aphasic patient who has difficulties producing complex onsets.

\begin{example}[VBR's complex onsets (\citealp{Buchwald2005}:79--80, his transcriptions)] \label{VBR}
\begin{tabular}{l l l}
a. & kəræb  & `crab'  \\
   & bəlid  & `bleed' \\
b. & kəwin  & `queen' \\
   & kəwoʊt & `quote' \\
c. & kut    & `cute'  \\
   & musɪk  & `music  \\ 
\end{tabular}
\end{example}

\noindent
VBR breaks up complex onsets with epenthesis, including those that consist of a consonant and a back onglide cluster (\ref{VBR}b). However, no epenthesis occurs between a consonant and a front onglide; rather, the glide is absent (\ref{VBR}c). The failure of the front onglide to pattern with other consonant clusters suggests once again that the glide is part of the nucleus.

%One potential problem with this account is noted by \citet{Kaye1996}, who obseres while [juː] may follow any single tautosyllabic consonant, it never follows branching onsets unless they consist of [s] and a single consonant. 
%This is the only sign that [juː] shows an affiliation for the onset.

\subsubsection{[w] onglides}

The selective properties of the back onglide [w] contrast sharply with those of the front onglide, and I assume that it is assigned to the onset. Whereas the front onglide shows only limited selectivity for preceding tautosyllabic consonants \citep{Davis1995,Kaye1996}, the back onglide [w] is rarely preceded by tautosyllabic consonants other than [k] (e.g., \emph{tran}[kw]\emph{il}). Unlike the front glide, syllable-initial [kw] may be followed by nearly any vowel \citep[][161]{Davis1995}. Unlike [juː], onglide [w] followed by a vowel does not pattern together in Pig Latin \citep[][166]{Davis1995}. And onsets followed by [w] pattern with other complex onsets in undergoing epenthesis in the patholical speech of VBR, discussed immediately above.

%Whereas [Cjuː] syllables attracts stress, [kwV] syllables do not \citep[][162f.]{Davis1995}.

\subsubsection{Post-vocalic \emph{r}}

Both CELEX and the dictionary used by \citet{Pierrehumbert1994} in her study transcribe Received Pronunciation, in which word-medial post-vocalic \emph{r} has been lost. In \emph{r}-full dialects, there is reason to believe that post-vocalic \emph{r} is in fact nuclear, and therefore its presence or absence is irrelevant to the constraints on syllable contact clusters. Many vowel contrasts are suspended before \emph{r} \citep[][255]{Harris1994}; for example, \emph{fern}, \emph{fir}, and \emph{fur} lack the contrasts found in \emph{pet}, \emph{pit}, and \emph{putt}. Further evidence for the nuclear status of post-vocalic \emph{r} comes from variable phonological processes in which post-vocalic \emph{r} patterns differently than internal coda consonants. \citeauthor{Harris1994} reports that a variable process of /t/-\textsc{Glottalization} in many dialects of British English is blocked when /t/ is preceded by any consonant except post-vocalic \emph{r}.

\begin{example}[/t/-\textsc{Glottalization} in British English (after \citealp{Harris1994}:195, 258)]
\begin{tabular}{l l l@{} l}
a. & fis[t]   & * & fis[ʔ]   \\
   & mis[t]er & * & mis[ʔ]er \\
b. & par[t]   &   & par[ʔ]   \\
   & car[t]on &   & car[ʔ]on \\
\end{tabular}
\end{example}

\noindent Similarly, while /t, d/ delete in word-final position when immediately preceded by a consonant, including sonorants /n, l/, as in (\ref{td}a), deletion of /t, d/ after post-vocalic \emph{r} in American English is ``rare or nonexistent'' \citep[][8]{Guy1980}.

\begin{example}[/t, d/-\textsc{Deletion}in American English] \label{td}
\begin{tabular}{l l l@{} l}
a. & be[lt]  &   & be[l]  \\
   & me[nd]  &   & me[n]  \\
b. & sh[ɚt]  & * & sh[ɚ] \\
   & c[ɚd]   & * & c[ɚ]  \\
\end{tabular}
\end{example}

%Finally, \citet[][251]{Fromkin1973} also presents evidence that post-vocalic \emph{r} may behave as if nuclear in speech errors.

\section{Evaluation}

% what is being evaluated
% summary

\subsection{Static and derived constraints}

With this database of English syllable contact clusters it is now possible to evaluate constraints on sequences of underlying phonemes.

Here, the metric used is simply

The evaluation

The foremost evidence for

The evaluation used here is simply

Having described the database of English syllable contact clusters,

I now consider constraints on sequences of URs as predictors of attested and unattested contact clusters. First, I describe the statistical technique used, then apply this technique to both phonologically derived constraints and a set of ``static'' phonotactic constraints proposed by \citet{Pierrehumbert1994}.

\subsubsection{Statistical technique}
\label{stattech}

Before the dawn of generative phonology, many linguists concerned themselves with documenting co-occurrence in various languages. \citet[][28]{Vogt1954} declares that a ``phonemic description of a language \ldots should also comprise a description of the phoneme combinations that occur or may be expected to occur in the language''. Studies in this tradition include the discussions of consononantal co-occurrence in Arabic and Javanese by \citet{Greenberg1950} and \citet{Uhlenbeck1950}, respectively. \citeauthor{Uhlenbeck1950}, for instance, puts forth co-occurrence dispreferences, rather than exceptionless generalizations of the sort given by \citet{Greenberg1950}.

\citet{McCarthy1988}

Years later, \citet{Mester1988} revisits \citeauthor{Uhlenbeck1950}'s study and proposes a statistical technique, based on the chi-square test, to distinguish between accidental and structural lexical dispreferences in the Javanese lexicon.

\citeauthor{Mester1988}'s technique has been adopted by many subsequent studies (e.g., \citealt{Padgett1992,Padgett1995} on Russian; \citealt{Pierrehumbert1993}, \citealt{Frisch1996}, and \citealt{Frisch2004} on Arabic; \citealt{Berkley1994b,Berkley1994a,Berkley2000}, \citealt{Dmitrieva2008a}, \citealt{Dmitrieva2008b} on English). The null hypothesis $H_0$ is that some sequence occurs at no different a rate than the researcher expects, and the alternative hypothesis $H_1$ is that it is underrepresented. Given an observed frequency $O$ and also a frequency $E$ expected under the null hypothesis, the test is as follows.

\begin{example}[null hypothesis testing]
$\displaystyle \textrm{Reject } H_0 \textrm{ iff } \frac{(O E) ^ 2}{E} >
χ^2$
\end{example}

%\footnote{This is the one-tailed version of the test.}

In this study, I use the \citet{Fisher1934} exact test, which is isomorphic to this chi-square test. Compared to the chi-square test, the Fisher exact test $p$-value is somewhat more difficult to compute, but both are automated by modern statistical packages and can be computed rapidly. The major advantage of the Fisher test over the chi-square test is that it is appropriate for small samples or rare events, whereas the chi-square test depends on an approximation which is only exact in the limit (i.e., with a hypothetically infinite sample), and therefore is not appropriate for small samples \citep[see][]{Gorman2012a}.
%\citep[e.g.,][142]{Mester1988}.

On the other end of the spectrum, the Monte Carlo statistical procedure developed by \citet{Kessler2001} and applied to co-occurrence statistics by \citet{Martin2007,Martin2011} and \citet{Brown2010} does not scale to large samples. The Monte Carlo procedure is isomorphic to the one-tailed Fisher Exact test, and involves comparing observed co-occurrence counts to those produced by randomly generated samples.  This requires the researcher to generate random permutations, something which is particularly difficult for large samples, as I will explain; disinterested readers may wish to skip the following paragraph.

Any permutation of a sequence $L$ can be represented as a sequence $S$ of the same length, where the value of $S_i$ corresponds to the permuted position of $L_i$. Generating random permutations requires an unbiased method to generate all $N!$ of these lists with equal probability. Unfortunately, this is near impossible for $N$s of moderate size with current computational resources. Any pseudo-random number generator can be characterized by its ``period'', the number of pseduo-random numbers it can emit before repeating itself; this is also the upper limit for the number of random permutations that can be generated. The pseudo-random number generator used by both \citeauthor{Martin2011} and \citeauthor{Brown2010} has a period of approximately $2^{48}$, a number which turns out to be smaller than $17!$. For the large samples, sometimes in the thousands, used by these researchers, virtually all the possible permutations can never be generated by this method. As a consequence, the shuffling procedure is far from random and introducing the possibility that the statistical test is biased in unpredictable ways.

As an example of the Fisher exact test as it is applied here, consider some of the Javanese co-occurrence facts considered by \citet{Mester1988}. \citeauthor{Mester1988} is at pains to show that there is a gradient dispreference for first and second root consonants which share the same major place. However, perfectly identical first and second consonants are exempt. Since

% APPLY THIS TO JAVANESE

\begin{example}[Javanese root co-occurrence] %(\citealp[][264]{Uhlenbeck1950}, \citealp[][139]{Mester1988})]
\begin{tabular}{l r r r r}
\toprule
           & attested & unattested & saturation & $p$-value \\
\midrule
conforming & 134      & 12         & 0.918      & \multirow{2}{*}{6.095\e{-06}} \\
violating  & 21       & 15         & 0.583 \\
\bottomrule
\end{tabular}
\end{example}

\noindent
It is evident that roots with identical labial stops are far more common than those in which the roots disagree in voicing.

This data is input to a chi-square test, which reports that this highly skewed distribution is unlikely to be due to chance

That this is highly unlikely to be due to chance is apparent both from the chi-square ($\chi^2 = 25.572$, two-tailed $p = 4.3$\e{-07}) and Fisher exaact tests ($p =  6.095$\e{-06}).

I use two-tailed statistical tests throughout. 
Under a one-tailed test, the alternative hypothesis is underrepresentation, whereas a two-tailed test also allows the possibility that a class of clusters would be overrepresented, i.e., more frequent than predicted by the null hypothesis. As

\citet{Pierrehumbert1994}

It must be stressed that under current assumptions, this is independent of whether predictable prosodic representations (particularly coda and onset identity) are part of the memorized underlying representation \citep[e.g.,][]{Vaux2003}.
It is an undeniable fact that URs need to be 
The derivative nature of 
\citet{Ito1989a,Noske1992}

least some researchers (i.e., \citealt{Mester1988})

 \citep[e.g.,]{Brown2010} have argued for phonotactic constraints giving rise to overrepresentation, so I admit this possibility here and use two-tailed tests throughout.

%\footnote{For example, \citet{Martin2007} applies this technique to a list of 4,758 noun-noun compounds. The number of permutations of a list of this length can be estimated using Stirling's approximation.

%\ex $ \lim_{n \rightarrow \infty} n! = e ^ {n \ln{n} - n} = 2 ^ {\log_2{e} (n \ln{n} - n)} $ \xe

%\noindent
%$4,758!$ is approximately $2 ^ {51260}$. The author is unaware of any system that provides 51,260 bits of entropy; and some popular programming languages, such as ANSI C, provide as few as 15 bits.}

%%% START OF SERIOUSNESS

I will first consider three well-known English phonological alternations, showing that these give rise to constraints on the sequence structure of morphs. Simple autosegmental formalizations of these processes are provided, and the Fisher exact test is used to demonstrate their effect on the English lexicon.

\subsubsection{Obstruent Voice Assimilation}

English has a process of voice assimilation which is found in two inflectional suffixes. The suffix /-d/ forms regular pasts and denominal adjectives, and /-z/ forms regular noun plurals and genitives, and the 3sg. present active indicative verbs.\footnote{The URs /-d, -z/ are argued for by \citet[][282]{Hockett1958}, \citet[][210]{SPE}, \citet{Basboll1972}, \citet{Shibatani1972}, S. \citet[][]{Anderson1973a}, \citet[][102]{Pinker1988}, and \citet[][284f.]{Bakovic2005b}, and alternative views are presented by \citet[][210f.]{LANGUAGE}, \citet[][426]{Nida1948}, \citet{Luelsdorff1969}, \citet{Lightner1970}, \citet{Hoard1971}, \citet[]{Miner1975}, \citet{Zwicky1975}, \citet{Kiparsky1985}, and \citet[][135]{Borowsky1986}.}

\begin{example}
Voice assimilation in the past tense and noun plural:

\vspace{0.5\baselineskip}
\begin{tabular}{l l l l l l}
a. & nap & nap[t] \\
   & nab & nab[d] \\
b. & lap & lap[s] \\
   & lab & lab[z]  \\
\end{tabular}
\end{example}

\noindent
The voiced obstruent suffixes devoice when preceded by a voiceless obstruent.
This is formalized as a rule spreading the [\textsc{Voice}] specification rightward over adjacent obstruents.

\begin{example}
\textsc{Obstruent Voice Assimilation} (\emph{SPE}:178):

\xymatrix@R=24pt@C=24pt{
\txt{[α \textsc{Voice}]}\ar@{-}[d]\ar@{..}[dr] & \\
\txt{C}\ar@{-}[d]             & \txt{C}\ar@{-}[d] \\
\txt{[$+$\textsc{Obstruent}]} & \txt{[$+$\textsc{Obstruent}]}\\
}
\end{example}

\noindent
This rules out the possibility of surface clusters of obstruents with differing  [\textsc{Voice}] specifications. While the data make it clear \textsc{Obstruent Voice Assimilation} applies in morphologically derived environments, it is not obvious that it applies to obstruent clusters in non-derived environments (e.g., obstruent clusters in the URs of morphs) as well. \citet{Pierrehumbert1994} does not mention voice assimilation in her study of gaps in English syllable contact clusters. \citet[][74f.]{Hammond1999a} notes the existence of words like \emph{a}[b.s]\emph{inth} and \emph{a}[s.b]\emph{estos}, which contain obstruent voice clusters disagreeing in [\textsc{Voice}], and takes this as evidence that obstruent voicing is restricted to derived environments.

Despite these exceptions, I propose that \textsc{Obstruent Voice Assimilation} plays a major role in shaping the English lexicon. Only English syllable contact clusters which contain two adjacent obstruents speak to the status of \textsc{Obstruent Voice Assimilation}; the remaining 160 possible clusters are ignored. A $2 \times 2$ contingency table is constructed by dividing the remaining possible clusters into those which agree on obstruent voicing and those which do not, and into those which are attested and those which are not.

\begin{example}[Lexical effects of \textsc{Obstruent Voice Assimilation}]
\vspace{0.5\baselineskip}
\begin{tabular}{l r r r r}
\toprule
           & attested & unattested & saturation & $p$-value \\
\midrule
conforming & 80 & 370 & 0.178 & \multirow{2}{*}{1.1\e{-11}}\\
violating  &  6 & 264 & 0.022 \\
\bottomrule
\end{tabular}
\end{example}

\noindent 
18\% of the possible clusters containing adjacent obstruents with the same [\textsc{Voice}] specification are found, whereas only 2\%  of those which disagree in [\textsc{Voice}], such as in \emph{a}[b.s]\emph{inth} and \emph{a}[s.b]\emph{estos}, CELEX also contains \emph{jo}[d.p]\emph{urs}, \emph{pi}[ntʃ.b]\emph{eck}, \emph{sa}[k.b]\emph{ut}, and \emph{ja}[k.d]\emph{aw}. As the $p$-value shows, the rarity of these disagreeing clusters is unlikely to be due to chance.

I propose that this pattern should be attributed directly to \textsc{Obstruent Voice Assimilation} applying to non-derived environments, preventing learners from positing URs in which adjacent obstruents do not agree in voice. There is no duplication between constraints on sequence structure and surface forms if both are derived from the application of the phonological rule (\citealt[][401f.]{Stanley1967}, \citealt[][382]{SPE}). The handful of exceptions to this generalization can be marked with appropriate exception diacritics. Another possibility is suggested by recent psycholinguistic results. \citet{Mattys2001b} find that 9-month old infants are sensitive to the contrast between [vt], a cluster of obstruents disagreeing in [\textsc{Voice}] which is not found word-medially, and [ft], which conforms to the above generalization and is found medially (e.g. \emph{a}[f.t]\emph{er}). This ability to use phonotactic violations to segment continuous speech persists into adulthood \citep{McQueen1998}, suggesting that ``hetero-voiced'' obstruent clusters may be analyzed as semantically opaque compounds by native speakers (e.g., \emph{ja}[k\#d]\emph{aw}). 
%\footnote{An analogous account is the proposal that readjustment rules insert morphological boundaries for the purposes of assigning stress (\emph{SPE}:94).}

\subsubsection{Coda nasal place assimilation}
\label{cnpasection}

Above, it was assumed that the velar nasal is derived from underlying /n/. 
This analysis is not purely allophonic, as nasal place assimilation alternations target the final consonant of the \emph{im-}/\emph{in-} prefix. In (\ref{npa}ab), the prefix-final consonant takes on the same major place of articulation as the following obstruent, and in (\ref{npa}c), where the root is vowel-initial, the prefix-final consonant is realized as a default /n/.

\begin{example}
\label{npa}
Nasal place assimilation in Latinate prefixes:

\vspace{0.5\baselineskip}
\begin{tabular}{l l l l}
%a. & possible & i[m.p]ossible & & balance & i[m.b]alance \\
%b. & tangible & i[n.t]angible & & decent  & i[n.d]ecent  \\
%c. & elegant  & i[n.]elegant  & & ability & i[n.]ability \\
a. & possible & i[m.p]ossible \\
   & balance  & i[m.b]alance  \\
b. & tangible & i[n.t]angible \\
   & decent   & i[n.d]ecent   \\
c. & elegant  & i[n]elegant  \\
   & ability  & i[n]ability  \\
\end{tabular} 
\end{example}

\noindent
The shape of the prefix before other obstruents is somewhat more complex. 
%Before /f/, CELEX transcribes the nasal as either labial or alveolar, the latter being very rare in simplex words. 
%There are also additional complexities regarding the shape of the prefix when attached to /k, g/-initial roots. 
\citet[][62]{Halle1985a} and \citet[][90]{Borowsky1986} report that coda nasals assimilate to [ŋ] before dorsal consonants, but this assimilation is blocked by stress on the following syllable, citing contrasts like the one between \emph{í}[ŋ.k]\emph{ubate} and \emph{i}[n.k]\emph{lúde}. The prefix \emph{com-}/\emph{con-}, which patterns with \emph{im-}/\emph{in-} in showing assimilation (e.g., \emph{co}[m.b]\emph{ine}), participates in a similar contrast, between \emph{có}[ŋ.g]\emph{ress} and \emph{co}[n.g]\emph{réssional}. As \citeauthor{Borowsky1986} notes, stress does not block assimilation in simplex words (e.g., \emph{a}[ŋ.g]\emph{óra}), a fact which demonstrates the necessity of a partially morphological explanation for the blocking of assimliation in words like \emph{co}[n.g]\emph{réssional}. Regardless of these complexities, there is sufficient evidence to posit assimilation of the \textsc{Place} features of an obstruent to a preceding nasal consonant.

\begin{example}
\textsc{Coda Nasal Place Assimilation} (\emph{SPE}:85):

\xymatrix@R=24pt@C=24pt{
                          & \txt{\textsc{Place}}\ar@{-}[d]\ar@{..}[dl] \\
\txt{C}\ar@{-}[d]         & \txt{C}\ar@{-}[d] \\
\txt{[$+$\textsc{Nasal}]} & \txt{[$+$\textsc{Obstruent}]} \\
}
\end{example}

\citet[][175]{Pierrehumbert1994} observes that in simplex English words, ``nasal-stop sequences agree in labiality''; presumably, this generalization is limited to [\textsc{Labial}] because \citeauthor{Pierrehumbert1994} also adopts the assumption that the velar nasal is derived from /n/. However, \citeauthor{Pierrehumbert1994} does not propose any connection between the alternation and the generalization that nasal consonants in URs tend to agree with the following consonant. While clusters like \emph{pi}[m.p]\emph{le}, \emph{sta}[n.z]\emph{a}, and \emph{mo}[ŋ.k]\emph{ey} are the norm, exceptions like \emph{pli}[m.s]\emph{oll}, \emph{da}[m.z]\emph{el}, \emph{scri}[m.ʃ]\emph{aw}, and \emph{ra}[m.k]\emph{in} are found.

In all, there are only 80 possible nasal-obstruent syllable contact clusters, the majority of which are attested. In addition, there are a small number of words which contain clusters like \emph{oi}[nt.m]\emph{ent}, in which a nasal-obstruent cluster are both parsed into a complex medial coda, but in this position, place agreement is exceptionless, and thus these clusters are ignored below.

\begin{example}[Lexical effects of \textsc{Coda Nasal Place Assimilation}]
\begin{tabular}{l r r r r}
\toprule
           & attested & unattested & saturation & $p$-value \\
\midrule
conforming & 41 & 11 & 0.788 & \multirow{2}{*}{2.6\e{-05}}\\
violating  & 8  & 20 & 0.286 \\
\bottomrule
\end{tabular}
\end{example}

\noindent Clusters which conform to this generalization are far more likely to be attested than those that violate it, and this is highly unlikely to be due to chance.

There is some evidence that even young infants show a preference for nonce words that conform to this generalization. %\citet{Jusczyk2003}
\citet{Davidson2004}, \citet{Mattys1999}, and \citet{Jusczyk2002} play nonce words containing nasal-obstruent clusters to 4.5, 9, and 10-month-old infants (respectively) acquiring English, All three studies find that infants prefer to listen to stimuli which conform to the above generalization (e.g., \emph{u}[m.b]\emph{o}) over those which do not (e.g., \emph{u}[n.b]\emph{o}). \citet{Wright1975} taught several nonce words to a group of adolescents, and asked them to repeat the words to each other in a game of ``Telephone''. In these nonse words, nasals do not agree in \textsc{Place} with the following obstruent in these nonsense words, and \citeauthor{Wright1975} observes that after a few rounds of the game, the nonce words had been adapted to conform to \textsc{Coda Nasal Place Assimilation}.

\begin{example}[Artificial language adaptations (\citealt{Wright1975}:394, his transcriptions)]
\begin{tabular}{l l l l l l}
a. & [gownp] & > & [gump] \\
b. & [ǰumg]  & > & [ǰúŋgə] \\
c. & [ðʌŋd]  & > & [tɔŋg] \\
\end{tabular}
\end{example}

\noindent If this game is akin to the process of loanword adaptation, it is possible that various extragrammatical pressures are at play \citep[e.g.,][]{Halle1998b,Dupoux1999,Ussishkin2003,Peperkamp2008}. Yet, the independent evidence for a phonological process of nasal place assimilation provides the most parsiminious account of the adaptations seen in \citeauthor{Wright1975}'s study. J. \citet{Myers1993} also shows that \textsc{Coda Nasal Place Assimilation} is fed by speech errors:

\begin{example}[Speech errors and \textsc{Coda Nasal Place Assimilation} (\citealp{Myers1993}:228)]
\begin{tabular}{l l l}
   & \emph{error}  & \emph{target} \\
a. & primps        & prints        \\
b. & rand orker    & rank order    \\
%  & [sprɪg]time for [hɪ̃ntlɘr] & springtime for Hitler \\
c. & po[ŋ]cutation & computation   \\
\end{tabular} 
\end{example}

%\citet{Hay2004a} demonstrates that [n.p, mθ] clusters, which are not found in English, are in fact rated better than many attested clusters.
%report that 10-month-old infants prefer to listen to nonsense words which conform to \textsc{Coda Nasal Place Assimilation} (e.g., \emph{u}[m.b]\emph{o}) compared to those which do not (e.g., \emph{u}[n.b]\emph{o}).

\subsubsection{Degemination}

A final well-known alternation in English is the simplification of geminates derived by certain suffixes. For instance, the suffix \emph{-ly} rarely forms a geminate when attaching to root-final /l/, whether forming denominal adjectives (e.g., \emph{sme}[l]\emph{y}) or deadjectival adverbs (e.g., \emph{norma}[l]\emph{y}). A small class of English verbs take a /-t/ suffix in the past; as this suffix does not trigger the \textsc{Epenthesis} found in the regular past (e.g., \citealt{Bakovic2005b}, \citealt{Fruehwald2011}), \textsc{Degemination} applies.

\begin{example}[Degemination in /-t/ pasts (\citealp{Halle1985a}:105, \citealp{Myers1987}:492)]
\begin{tabular}{l l l l l}
a. & burn    & burn[t] \\
   & spill   & spil[t] \\
b. & ben[d]  & ben[t]  \\
   & buil[d] & buil[t] \\
\end{tabular}
\end{example}

More complicated cases of degemination are found in English prefix morphology and described in \emph{SPE} (p.~148, 219f.) and \citealt[][]{Borowsky1986} (p. 102f.), and are not reviewed here.

%\ex \textsc{Coda Nasal Place Assimilation} feeds \textsc{Degemination} \citep[][116]{Borowsky1986}: \\
%\begin{tabular}{l l l l}
%a. & mature    & i[m]ature    & (cf. \emph{i}[m.b]\emph{alance})   \\
%b. & numerable & i[n]umerable & (cf. \emph{i}[n.d]\emph{ependent}) \\
%\end{tabular}
%\xe

The rule deletes the first of two adjacent consonants agreeing in \textsc{Place} and manner features, and is formalized using quantification over features \citep{Reiss2003b}.
% cf. \citep{Berent2012}

\begin{example}[\textsc{Degemination} (\emph{SPE}:253)]
\xymatrix@R=24pt@C=24pt{
\textsc{C}_1       & \longrightarrow & \emptyset & / & A~\gap\gap~A & \textsc{C}_2 \\
}
Condition: $\forall F_i \in \{\textsc{Place}, \textsc{Sonorant}, \textsc{Nasal}, \textsc{Continuant}, \textsc{Lateral}\}$ . $(F_i)_1 = (F_i)_2$
\end{example}

\noindent I assume that geminates crossing a syllable boundary are ``false geminates'', meaning that they are adjacent similar segments rather than a single feature matrix linked to two timing slots. Under the alternative account in which the absence of geminates is an inventory restriction (i.e., a \textsc{Segment Structure Condition}), an undesirable duplication of phonological derived surface constraints and constraints on URs would be unavoidable.

Neither identical consonants, nor those which differ only in voice (e.g., *[p.b]) occurs in simplex words in CELEX, and this absence is highly unlikely to be due to chance.

\begin{example}[Lexical effects of \textsc{Degemination}] 
\begin{tabular}{l r r r r}
\toprule
           & attested & unattested & saturation & $p$-value                   \\
\midrule
conforming & 157      & 596        & 0.208      & \multirow{2}{*}{7.2\e{-09}} \\
violating  & 0        &  87        & 0.000                                    \\
\bottomrule
\end{tabular}
\end{example}

%\footnote{Note that it is not the case that the English lexicon as envisioned by \emph{SPE} is free of underlying geminates; \citeauthor{SPE} use underlying geminates to attract stress to a preceding nucleus (p.~148--151). In \emph{SPE}, the underlying geminate-singleton contrast is absolutely neutralized, but has consequences for surface stress. In addition to the usual objections to absolute neutralization (i.e., that it is a learning conundrum mascarading as a phonological solution) that emerged in the wake of the publication of \emph{SPE}, the phonological particulars of this account have also come under attack \citep[e.g.,][]{Ross1972a}. It would not seem tenable to use underlying geminates to attract stress even if absolutely neutralized contrasts were admitted.} Unsurprisingly, this is unlikely to be due to chance.

\subsubsection{Dorsal/non-coronal clusters}

\citet{Pierrehumbert1994} surveys the English syllable contact cluster inventory, and discerns the effect of several ``static'' constraints, those which do not derive from neutralizing phonological rules.
